---
title: "Are Google Suggestions Sexist?"
author: "FD"
date: "7 June 2016"
output: html_document
---

Google's autocomplete feature sometimes yields disturbing results. When looking for female scientists, you may have noticed the appearance of "husband" in the list of suggestions, as reported recently [on Twitter](https://twitter.com/pollyp1/status/733636162263351296). But does this only happen to female scientists, or instead to any public figure?

To answer this question, I gathered names of high profile people with different professions (scientists, tennis players, Hollywood actors), wrote a script to fetch the list of Google suggestions, and compared the proportion of people for whom "husband" and/or "wife" are suggested as a function of their profession (see the [Methods](#Methods) section for more details). 

 ```{r echo=FALSE, include=FALSE}
# Clean the memory
rm(list=ls())

# For the report
#install.packages("rmarkdown")
library(rmarkdown, quietly=TRUE)


#===============================#
# OBTAIN AND COMBINE DATA		#
#===============================#

# Combine data for a given "category"
getdata <- function(catname){
	tmp <- read.csv(paste(catname, ".csv", sep=""), header=FALSE, stringsAsFactors=FALSE)
	# Find duplicates
	dupl <- duplicated(tmp$V1)

	wife <- scan(paste(catname, "_wife.txt", sep=""))
	husband <- scan(paste(catname, "_husband.txt", sep=""))
	n <- length(wife) - sum(dupl)
	
	# Combine data, removing duplicates
	t1 <- cbind(tmp[!dupl,], wife=1-wife[!dupl], husband=1-husband[!dupl])
	t2 <- cbind(t1, wh=1*((t1$wife==1)|(t1$husband==1)), sex=1*(t1[,2]=="F")) # F=1, M=0
	names(t2)[1:3] <- c("name", "MF", "cat")
	return(t2)
}

# Do it for all our datasets
t.aaas <- getdata("aaas")
t.ted <- getdata("ted")
t.tennis <- getdata("tennis")
t.actors <- getdata("actors")
t.hhmi <- getdata("hhmi")

# Now we want to combine the Science data (ted, hhmi, aaas)
 # There are people who belong to different categories: we need to merge them
 # So first, find them
  tmpTA <- rbind(t.ted, t.aaas) # Combine TED and AAAS datasets
  duTA <- duplicated(tmpTA[, 1]) # Find replicated names
  namesTA <- tmpTA[duTA, 1] # Get the corresponding names
  #print(namesTA) # Print replicates
  tmp <- tmpTA[!duTA,] # Remove replicates

  tmpTH <- rbind(tmp, t.hhmi) # Add HHMI data
  duTH <- duplicated(tmpTH$name) # Find replicated names
  namesTH <- tmpTH[duTH, 1] # Get the corresponding names 
  #print(namesTH) # Print replicates
  tmp <- tmpTH[!duTH,] # Remove replicates

 # Now we need to add the removed info again!
  t.science <- data.frame(cbind(tmp[, c("name", "sex")], "science", tmp[, c("wife", "husband", "wh")], 1*(tmp[3]=="AAAS"), 1*(tmp[3]=="HHMI"), 1*(tmp[3]=="TED")))
  names(t.science) <- c("name", "sex", "cat", "wife", "husband", "wh", "AAAS", "HHMI", "TED")
  # Add removed AAAS data
  for (dTA in namesTA){
	t.science[t.science$name==dTA, "AAAS"] <- 1
  }
# Add removed HHMI data
  for (dTH in namesTH){
	t.science[t.science$name==dTH, "HHMI"] <- 1
  }

manualcontrol <- FALSE
if(manualcontrol){
	# Manual control of remaining potential duplicates
	library(stringr)
	# Extract last word in name (mostly last name)
	nm <- unname(vapply(t.science$name, function(l) word(l, -1), FUN.VALUE="c(1)"))
	# Sort by alphabetical order
	onm <- nm[order(nm)]
	# Sort the entire table by alphabetical order
	temp <- t.science[order(nm),]
	# Find duplicates in last names (all values, hence the | and fromLast) and print them
	temp[duplicated(onm) | duplicated(onm, fromLast=TRUE),]
	# Then inspect the table and potentially correct the files if typoes
	# But this is not necessary anymore since it's already been done.
}
	
#===============================#
# STATS AND OTHER CALCULATIONS  #
#===============================#

# Count the number of F/M, wife+husband
countsuggestions <- function(cat){
	tab <- get(paste("t.", cat, sep=""))
	nF <- sum(tab$sex==1)
	nM <- sum(tab$sex==0)
	nFwh <- sum(tab$sex==1 & tab$wh==1)
	nMwh <- sum(tab$sex==0 & tab$wh==1)
	out <- c(nF=nF, nM=nM, nFwh=nFwh, nMwh=nMwh, pFwh=nFwh/nF, pMwh=nMwh/nM)
}

cats <- list("actors", "tennis", "science")
cts <- t(sapply(cats, countsuggestions))
x <- data.frame(cts)
xx <- cbind(data.frame(cat=unlist(cats)), x)
names(xx) <- c("cat", dimnames(cts)[[2]])

 ```

# Results

## Comparing professions
```{r, echo=FALSE}
# PLOTS
colM <- rgb(0, 150, 150, 100, maxColorValue = 255)
colF <- rgb(249, 80, 0, maxColorValue = 255)

cols <- c("sex", "wh", "cat")
#t.all <- rbind(t.tennis[, cols], t.actors[, cols], t.science[, cols])
#counts <- table(t.all$wh, t.all$cat)
par(las=1)
barplot(t(as.matrix(xx[, c("pMwh", "pFwh")])), #
        beside=TRUE, col=c(colM, colF), names.arg=c("actors", "tennis players", "scientists"), ylim=c(0,1), main="Proportion of individuals for whom \n'wife' and/or 'husband' appear as suggestion in a Google search", cex.main=0.85, legend = c("Men", "Women"))
```

### Actors
There is basically no difference between male (`r sprintf("%.2f",  xx$pMwh[xx$cat=="actors"])`) and female (`r sprintf("%.2f", xx$pFwh[xx$cat=="actors"])`) actors in terms of frequency of "wife" or "husband" suggestions by Google:
```{r}
summary(glm(formula = wh ~ sex, data=t.actors, family = binomial))
```

### Tennis players
There is a strong difference between male and female tennis players, but maybe not the one you'd expect: "husband" or "wife" appear in Google's suggestions for `r xx$nFwh[xx$cat=="tennis"]` women out of the `r sum(t.tennis$sex)` in the dataset, while they appear for `r xx$nMwh[xx$cat=="tennis"]` men (out of `r sum(1-t.tennis$sex)`). One reason may be that, for an equivalent ranking, the popularity of female tennis players is lower than the popularity of male tennis players. 
```{r}
summary(glm(formula = wh ~ sex, data=t.tennis, family = binomial))
```

### Scientists
If we now turn to the scientists dataset, we first observe that the frequency at which the "wife" and "husband" suggestions appear is much lower (`r mean(t.science$wh)` for men and women combined) than for the actors dataset for instance. What about the difference between men and women in this dataset?
```{r}
summary(glm(formula = wh ~ sex, data=t.science, family = binomial))
```
We do not detect an significant difference in this dataset. 

## Further dissecting the "science" dataset

The suggestions highlighted on [Twitter](https://twitter.com/pollyp1/status/733636162263351296) were mainly about HHMI scientists, so let's focus on this subset of our science dataset for the moment: 
```{r}
summary(glm(formula = wh ~ sex, data=t.science[t.science$HHMI==1,], family = binomial))
```

There is indeed an effect of sex on the proportion of "wife" and "husband" in Google Suggestions for HHMI scientists. But if we look closer, we realize that out of the `r sum(t.science[t.science$sex==1 & t.science$HHMI==1, "wh"])` HHMI female scientists for whom "husband" or "wife" appears in Google's suggestions, `r sum(t.science[t.science$sex==1 & t.science$HHMI==1 & t.science$TED==1, "wh"])` have given TED talks, out of the `r length(t.science[t.science$sex==1 & t.science$HHMI==1 & t.science$TED==1, "wh"])` female HHMI scientists who have given TED talks. Out of the `r sum(t.science[t.science$sex==0 & t.science$HHMI==1, "wh"])` male HHMI scientists for whom "husband" or "wife" appears in Google's suggestions, `r sum(t.science[t.science$sex==0 & t.science$HHMI==1 & t.science$TED==1, "wh"])` have given TED talks, out of the `r length(t.science[t.science$sex==0 & t.science$HHMI==1 & t.science$TED==1, "wh"])` male HHMI scientists who have given TED talks.

It may be the higher proportion of TED speakers among female HHMI scientists that is driving the result. Here is a summary of the different sample sizes among HHMI scientists:
```{r echo=FALSE}
tab <- tapply(t.science[t.science$HHMI==1, "wh"], list(sex=t.science[t.science$HHMI==1,"sex"], ted=t.science[t.science$HHMI==1,"TED"]), length)
rownames(tab) <- c("M", "W")
colnames(tab) <- c("no", "yes")
tab
```
Now adding participation to a TED talk in the analysis of this HHMI dataset,

```{r}
summary(glm(formula = wh ~ sex + TED, data=t.science[t.science$HHMI==1, ], family = binomial))
```

```{r}
summary(glm(formula = wh ~ sex * TED, data=t.science[t.science$HHMI==1, ], family = binomial))
```

We will hence go back to the entire science dataset, which includes AAAS (American Academy of Arts and Sciences) members (Section II.4) and all TED speakers of the science section. 

With this entire science dataset, we can explore the interactions between sex and the fact of having given a TED talk (i.e., having been given a wide exposure) on the probability of having "wife" or "husband" appear as a Google suggestion:

```{r}
summary(glm(formula = wh ~ sex+TED, data=t.science[t.science$HHMI==1 | t.science$AAAS==1, ], family = binomial))
```

This analysis suggests that the TED effect is strongest. 

```{r echo=FALSE}
darken <- function(color, factor=1.4){
    rgb(t(col2rgb(color)/factor), maxColorValue=255)
}

colMT <- darken(colM)
colFT <- darken(colF)
dissectscience <- tapply(t.science$wh, list(TED=t.science$TED, sex=t.science$sex), sum)
dissectscience <- dissectscience / matrix(rep(c(xx[xx$cat=="science", "nM"], xx[xx$cat=="science", "nF"]), 2), byrow=TRUE, ncol=2)
dmod <- cbind(c(dissectscience[,1], 0, 0), c(0, 0, dissectscience[,2])) # Trick to make R believe that there are four categories, then we can tune the colors as we wish
par(las=1)
#barplot(dmod, beside=FALSE, col=matrix(c(colM, colM, colF, colF), ncol=2), names.arg=c("Men", "Women"), xlim=c(0, 0.1), legend=c("1", "2", "3", "4"))
barplot(dmod, beside=FALSE, col=matrix(c(colMT, colM, colFT, colF), ncol=2), border=TRUE, legend=c("Men, TED", "other men", "Women, TED", "Other women"), ylim=c(0, 0.1), names.arg=c("Men", "Women"), main="Proportion of scientists for whom \n'wife' or 'husband' appear as a suggestion in a Google search", cex.main=0.85)
```

# Conclusion

Google's suggestions may not be that sexist against female scientists after all: the fact that "wife" and "husband" appear in the list of suggestions of a Google search may instead be a side-effect of being a public figure. 

# <a name="Methods" style="text-decoration:none;">Methods</a>
## Name collection
### Tennis players
I downloaded on 2016-06-06 the names of the 50 best-ranked male and female tennis players (according to the [ATP](http://www.atpworldtour.com/en/rankings/singles) and [WTA](http://www.wtatennis.com/singles-rankings) rankings). [(download the list of names)](tennis.csv).

### Actors
I downloaded on 2016-06-06 a [list of the 100 most valuable movie stars](http://www.vulture.com/2012/07/most-valuable-movie-stars.html), assigned sexes based on first names and pictures. The list contains `r sum(t.actors$sex)` women names and `r sum(1-t.actors$sex)` men names. [(download the list of names)](actors.csv).

### Scientists
I downloaded on 2016-06-06 three datasets: 

- The names of all AAAS members in the [Evolutionary and Population Biology and Ecology section](https://www.amacad.org/multimedia/pdfs/classlist.pdf) (`r sum(t.aaas$sex)` women, `r sum(1-t.aaas$sex)` men, [download the list of names](aaas.csv).);

- The names of all [HHMI scientists](https://www.hhmi.org/scientists/browse?kw=&sort_by=field_scientist_last_name&sort_order=ASC&items_per_page=100) (`r sum(t.hhmi$sex)` women, `r sum(1-t.hhmi$sex)` men, [download the list of names](hhmi.csv).);

- The names of all TED speakers in the [Science section](https://www.ted.com/talks?sort=popular&topics%5B%5D=science&language=en) (`r sum(t.ted$sex)` women, `r sum(1-t.ted$sex)` men, [download the list of names](ted.csv).).

Some names appear in more than one of these lists. I combined all names in a table, removed duplicates, and indicated with 0s and 1s the affiliations of each scientist, e.g.
```{r, echo=FALSE}
t.science[t.science$name=="Edward O. Wilson", ]
```
for a total of `r length(t.science$sex)` individual scientists (`r sum(t.science$sex)` women and `r sum(1-t.science$sex)` men). 

## Google Suggestions
I wrote a [bash script](script.sh) to do this step automatically. Google suggestions for each name in each list were downloaded via the <http://suggestqueries.google.com> webpage. For each set of suggestions, the words "wife" and "husband" were searched for.

Please note that although I made sure to specify `hl=en` in the query (i.e., setting the language to English), Google Suggestions may differ depending on geographic location. You can do this step again simply by typing 
```{sh eval=FALSE}
./script.sh fileprefix
```
in a terminal, where `fileprefix` is either `tennis`, `actors`, `aaas`, `hhmi`, or `ted`.

## Analysis
You just need to look at the [source code](report.Rmd) of this file! It is written in Rmarkdown, so the analysis code is contained in it. 

## Reproducibility
As mentioned above, the results will depend on where you run the analysis from, since Google suggestions depend on location. 
To redo everything on the already collected lists of names, simply type
```{sh eval=FALSE}
./make.sh
```
in a terminal (in the working directory of the project.)


  
    
    
    
    


    
    
    
