---
title: "Are Google Suggestions Sexist?"
author: "FD"
date: "7 June 2016"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Google's autocomplete feature sometimes yields disturbing results. When looking for female scientists, you may have noticed the appearance of "husband" in the list of suggestions, as reported recently [on Twitter](https://twitter.com/pollyp1/status/733636162263351296). But does this only happen to female scientists?

To answer this question, I gathered names of high profile people with different professions (scientists, tennis players, Hollywood actors), wrote a script to fetch the list of Google suggestions, and considered the proportions of people for whom "husband" and/or "wife" are suggested; I also looked for the word "married". 

**tl;dr:** It looks like the proportions of "wife" and/or "husband" are higher for the female scientists of our dataset (HHMI scientists as well as AAAS members in Class II: Biological Sciences) than for male scientists, but that they are also higher for scientists of our dataset who have given TED talks compared to those who have not. 

```{r echo=FALSE, include=FALSE}
# Clean the memory
rm(list=ls())

# For the report
#install.packages("rmarkdown")
library(rmarkdown, quietly=TRUE)


#===============================#
# OBTAIN AND COMBINE DATA       #
#===============================#

# Combine data for a given "category"
getdata <- function(catname){
	tmp <- read.csv(paste(catname, ".csv", sep=""), header=FALSE, stringsAsFactors=FALSE)
	# Find duplicates
	dupl <- duplicated(tmp$V1)

	wife <- scan(paste(catname, "_wife.txt", sep=""))
	husband <- scan(paste(catname, "_husband.txt", sep=""))
	married <- scan(paste(catname, "_married.txt", sep=""))
	n <- length(wife) - sum(dupl)
	
	# Combine data, removing duplicates
	t1 <- cbind(tmp[!dupl,], wife=1-wife[!dupl], husband=1-husband[!dupl], married=1-married[!dupl])
	# "1-" because "yes" is coded 0 in the wife and husband data.
	whcombined <- ((t1$wife==1)|(t1$husband==1))
	#whv <- vapply(whcombined, function(x) if(x) "yes" else "no", FUN.VALUE="a")
	t2 <- cbind(t1, wh=1*whcombined) 
	names(t2)[1:3] <- c("name", "sex", "cat")
	t2[t2$sex=="F","sex"] <- "W" # Changing the way women are encoded to change the alphabetical order btw F/M vs M/W
	return(t2)
}

# Do it for all our datasets
t.aaas <- getdata("aaas")
t.ted <- getdata("ted")
t.tennis <- getdata("tennis")
t.actors <- getdata("actors")
t.hhmi <- getdata("hhmi")

# Now we want to combine the Science data (ted, hhmi, aaas)
 # There are people who belong to different categories: we need to merge them
 # So first, find them
  tmpTA <- rbind(t.ted, t.aaas) # Combine TED and AAAS datasets
  duTA <- duplicated(tmpTA[, 1]) # Find replicated names
  namesTA <- tmpTA[duTA, 1] # Get the corresponding names
  #print(namesTA) # Print replicates
  tmp <- tmpTA[!duTA,] # Remove replicates

  tmpTH <- rbind(tmp, t.hhmi) # Add HHMI data
  duTH <- duplicated(tmpTH$name) # Find replicated names
  namesTH <- tmpTH[duTH, 1] # Get the corresponding names 
  #print(namesTH) # Print replicates
  tmp <- tmpTH[!duTH,] # Remove replicates

 # Now we need to add the removed info again!
  t.science <- data.frame(cbind(tmp[, c("name", "sex")], "science", #
                                tmp[, c("wife", "husband", "wh", "married")], #
                                unname(vapply(tmp[,3], function(x){if(x=="AAAS") "yes" else "no"}, FUN.VALUE="a")), #
                                unname(vapply(tmp[,3], function(x){if(x=="HHMI") "yes" else "no"}, FUN.VALUE="a")), #
                                unname(vapply(tmp[,3], function(x){if(x=="TED") "yes" else "no"}, FUN.VALUE="a")))#
                          )
  names(t.science) <- c("name", "sex", "cat", "wife", "husband", "wh", "married", "AAAS", "HHMI", "TED")
  # Add removed AAAS data
  for (dTA in namesTA){
	t.science[t.science$name==dTA, "AAAS"] <- "yes"
  }
# Add removed HHMI data
  for (dTH in namesTH){
	t.science[t.science$name==dTH, "HHMI"] <- "yes"
  }

manualcontrol <- FALSE
if(manualcontrol){
	# Manual control of remaining potential duplicates
	library(stringr)
	# Extract last word in name (mostly last name)
	nm <- unname(vapply(t.science$name, function(l) word(l, -1), FUN.VALUE="c(1)"))
	# Extract last word in name (mostly last name)
	fnm <- unname(vapply(t.science$name, function(l) word(l, 1), FUN.VALUE="c(1)"))
	# Sort by alphabetical order
  aord <- order(nm)
	onm <- nm[aord]
	# Sort the entire table by alphabetical order
	temp <- t.science[aord,]
	# Find duplicates in last names (all values, hence the | and fromLast) and print them
	duplnm <- temp[duplicated(onm) | duplicated(onm, fromLast=TRUE),]
	duplnm
	
	# Focus on lines with same first and last words
  ofnm <- fnm[aord]
  # Merge first and last names
  omerge <- vapply(seq_along(onm), function(i){paste(ofnm[i], onm[i], sep=" ")}, FUN.VALUE="a")
	# Find duplicates in last names (all values, hence the | and fromLast) and print them
	temp[(duplicated(omerge) | duplicated(omerge, fromLast=TRUE)),]

  cat('now inspect the abover table to find potential duplicates. \n')  
	# Then inspect the table and potentially correct the files if typoes
	# But this is not necessary anymore since it's already been done.
}
	
#===============================#
# STATS AND OTHER CALCULATIONS  #
#===============================#

# Count the number of F/M, wife+husband
countsuggestions <- function(cat){
	tab <- get(paste("t.", cat, sep=""))
	nF <- sum(tab$sex=="W")
	nM <- sum(tab$sex=="M")
	nFwh <- sum(tab$sex=="W" & tab$wh==1)
	nMwh <- sum(!tab$sex=="W" & tab$wh==1)
	nFm <- sum(tab$sex=="W" & tab$married==1)
	nMm <- sum(!tab$sex=="W" & tab$married==1)
	out <- c(nF=nF, nM=nM, #
	         nFwh=nFwh, nMwh=nMwh, pFwh=nFwh/nF, pMwh=nMwh/nM, #
	         nFm=nFm, nMm=nMm, pFm=nFm/nF, pMm=nMm/nM)
}

cats <- list("actors", "tennis", "science")
cts <- t(sapply(cats, countsuggestions))
x <- data.frame(cts)
xx <- cbind(data.frame(cat=unlist(cats)), x)
names(xx) <- c("cat", dimnames(cts)[[2]])
```

# Methods
## Name collection
### Tennis players
I downloaded on 2016-06-06 the names of the 50 best-ranked male and female tennis players (according to the [ATP](http://www.atpworldtour.com/en/rankings/singles) and [WTA](http://www.wtatennis.com/singles-rankings) rankings). [(download the list of names)](tennis.csv).

### Actors
I downloaded on 2016-06-06 a [list of the 100 most valuable movie stars](http://www.vulture.com/2012/07/most-valuable-movie-stars.html), assigned sexes based on first names and pictures. The list contains `r sum(t.actors$sex=="W")` women names and `r sum(t.actors$sex=="M")` men names. [(download the list of names)](actors.csv).

### Scientists
I downloaded on 2016-06-06 three datasets: 

- The names of all AAAS members in the [Biological Sciences Class (II)](https://www.amacad.org/multimedia/pdfs/classlist.pdf) (`r sum(t.aaas$sex=="W")` women, `r sum(t.aaas$sex=="M")` men, [download the list of names](aaas.csv).);

- The names of all [HHMI scientists](https://www.hhmi.org/scientists/browse?kw=&sort_by=field_scientist_last_name&sort_order=ASC&items_per_page=100) (`r sum(t.hhmi$sex=="W")` women, `r sum(t.hhmi$sex=="M")` men, [download the list of names](hhmi.csv).);

- The names of all TED speakers in the [Science section](https://www.ted.com/talks?sort=popular&topics%5B%5D=science&language=en) (`r sum(t.ted$sex=="W")` women, `r sum(t.ted$sex=="M")` men, [download the list of names](ted.csv).).

Sexes were assigned on the basis of first names and pictures (apologies if I made mistakes). 

Some names appear in more than one of these lists. I combined all names in a single table, removed duplicates, and indicated with "yes" and "no" the affiliations of each scientist, e.g.
```{r, echo=FALSE}
t.science[t.science$name=="Edward O. Wilson", ]
```
for a total of `r length(t.science$sex)` individual scientists (`r sum(t.science$sex=="W")` women and `r sum(t.science$sex=="M")` men). The `wh` column above combines the `wife` and `husband` information (`1` means that the words "wife" and/or "husband" appear in the suggestions). 

## Google Suggestions
I wrote a [bash script](script.sh) to do this step automatically. Google suggestions for each name in each list were downloaded via the <http://suggestqueries.google.com> webpage. For each set of suggestions, the words "wife", "husband", and also "married" were searched for.

Please note that although I made sure to specify `hl=en` in the query (i.e., setting the language to English), Google Suggestions may differ depending on geographic location. You can do this step again simply by typing 
```{sh eval=FALSE}
./script.sh fileprefix
```
in a terminal, where `fileprefix` is either `tennis`, `actors`, `aaas`, `hhmi`, or `ted`.

## Analysis
You just need to look at the [source code](report.Rmd) of this file! It is written in Rmarkdown, so the analysis code is contained in it. 

## Reproducibility
As mentioned above, the results will depend on where you run the analysis from, since Google suggestions depend on location. 
To redo everything on the already collected lists of names, simply type
```{sh eval=FALSE}
./make.sh
```
in a terminal (in the working directory of the project.)

# Results

## For different professions
```{r, echo=FALSE}
# PLOTS
colM <- rgb(0, 150, 150, 100, maxColorValue = 255)
colF <- rgb(249, 80, 0, maxColorValue = 255)

par(las=1)
barplot(t(as.matrix(xx[, c("pMwh", "pFwh")])), #
        beside=TRUE, col=c(colM, colF), names.arg=c("actors", "tennis players", "scientists"), ylim=c(0,1), main="Proportions, by gender, of individuals for whom \n'wife' and/or 'husband' appear as suggestion in a Google search", cex.main=0.85, legend = c("Men", "Women"))
```

### Actors
There is basically no difference between male (`r round(xx$pMwh[xx$cat=="actors"], 2)`) and female (`r round(xx$pFwh[xx$cat=="actors"], 2)`) actors in terms of frequency of "wife" or "husband" suggestions by Google:
```{r}
summary(glm(formula = wh ~ sex, data=t.actors, family = binomial))
```


### Tennis players
There is a strong difference between male and female tennis players, but maybe not the one you'd expect: "husband" or "wife" appear in Google's suggestions for `r xx$nFwh[xx$cat=="tennis"]` women out of the `r sum(t.tennis$sex=="W")` in the dataset, while they appear for `r xx$nMwh[xx$cat=="tennis"]` men (out of `r sum(t.tennis$sex=="M")`). One reason may be that, for an equivalent ranking, the popularity of female tennis players is lower than the popularity of male tennis players. 
```{r}
summary(glm(formula = wh ~ sex, data=t.tennis, family = binomial))
```

### Scientists
If we now turn to the scientists dataset, we first observe that the frequency at which the "wife" and "husband" suggestions appear is much lower (`r round(mean(t.science$wh), 2)` for men and women combined) than for the actors dataset for instance. What about the difference between men and women in this dataset?
```{r}
summary(glm(formula = wh ~ sex, data=t.science, family = binomial))
```

There is a significant effect of sex; the proportions of "wife" and/or "husband" are higher among Google suggestions for female scientists. 

## Further dissecting the "science" dataset

Some TED speakers are not scientists, but science journalists. So here, we will focus on the AAAS+HHMI datasets, keeping the information of whether the scientists have given TED talks. TED talks are considered here as a proxy of fame / public exposure (some TED talks have millions of views!). It is likely that the more well-known a scientist is, the more likely he/she is to be Googled... and maybe the more likely it is to see "wife" and/or "husband" appear in the suggestions.  

Here are the sizes of each categories of the AAAS+HHMI data:
```{r, echo=FALSE}
sciencesubset <- t.science[t.science$HHMI=="yes" | t.science$AAAS=="yes", ]
tapply(sciencesubset$wh, list(TED=sciencesubset$TED, sex=sciencesubset$sex), length)
```

### "wife" and/or "husband" for HHMI + AAAS scientists

Let's compare different models including sex and/or TED as factors:
```{r, echo=FALSE, include=FALSE}
#install.packages("MuMIn")
library(MuMIn, quiet=TRUE)
```

```{r}
model.hhmiaaas.wh.s <- glm(formula = wh ~ sex, data=sciencesubset, family = binomial)
model.hhmiaaas.wh.t <- glm(formula = wh ~ TED, data=sciencesubset, family = binomial)
model.hhmiaaas.wh.st <- glm(formula = wh ~ sex + TED, data=sciencesubset, family = binomial)
model.hhmiaaas.wh.i <- glm(formula = wh ~ sex * TED, data=sciencesubset, family = binomial)

# Adding new stuff for Nico
sciencesubset$Int <- as.factor(paste(sciencesubset$sex, sciencesubset$TED, sep="_"))
levels(sciencesubset$In) <- c("M", "M", levels(sciencesubset$In)[3], levels(sciencesubset$In)[4])

model.hhmiaaas.wh.nico <- glm(wh ~ Int, family = binomial, data = sciencesubset)

model.sel(model.hhmiaaas.wh.s, model.hhmiaaas.wh.t, model.hhmiaaas.wh.st, model.hhmiaaas.wh.i, model.hhmiaaas.wh.nico)
```

The best model seems to be the one that takes into account sex, TED talks, and the interaction between sex and TED talks:
```{r}
summary(model.hhmiaaas.wh.i)
```

Here are the numbers of AAAS and HHMI scientists for whom "wife" and/or "husband" appear as suggestions:
```{r echo=FALSE}
dissectscience <- tapply(sciencesubset$wh, list(TED=sciencesubset$TED, sex=sciencesubset$sex), sum)
dissectscience
```

In terms of proportions for each sex, this corresponds to
```{r echo=FALSE}
darken <- function(color, factor=1.4){
    rgb(t(col2rgb(color)/factor), maxColorValue=255)
}

colMT <- darken(colM)
colFT <- darken(colF)

dissectscience <- dissectscience / matrix(rep(c(sum(sciencesubset$sex==dimnames(dissectscience)$sex[1]), sum(sciencesubset$sex==dimnames(dissectscience)$sex[2])), 2), byrow=TRUE, ncol=2)
dissectscience <- dissectscience[c("yes", "no"), ] # Invert the two lines

dmod <- cbind(c(dissectscience[,1], 0, 0), c(0, 0, dissectscience[,2])) # Trick to make R believe that there are four categories, then we can tune the colors as we wish
par(las=1)
barplot(dmod, beside=FALSE, col=matrix(c(colMT, colM, colFT, colF), ncol=2), border=TRUE, legend=c("Men, TED", "other men", "Women, TED", "Other women"), ylim=c(0, 0.1), names.arg=c("Men", "Women"), main="Proportions, by gender, of AAAS and HHMI scientists for whom \n'wife' and/or 'husband' appear as a suggestion in a Google search", cex.main=0.85)
```

```{r echo=FALSE, include=FALSE, eval=FALSE}
png("fig.png", width=450, height=400)
par(mar=c(4, 7, 4, 6))
par(las=1)
barplot(dmod, beside=FALSE, col=matrix(c(colMT, colM, colFT, colF), ncol=2), border=TRUE, ylim=c(0, 0.1), names.arg=c("Men", "Women"), main="Proportions, by gender, of all AAAS members in Class II: Biological Sciences\n and HHMI scientists for whom 'wife' and/or 'husband' \nappear as a suggestion in a Google search", cex.main=1)
thecex <- 1.15
text(1.9, 0.00, "have given \nTED talks", pos=3, cex=thecex)
text(1.9, 0.03, "have NOT given \nTED talks", pos=3, cex=thecex)
text(0.7, 0.003, "have NOT given \nTED talks", pos=3, cex=thecex)
dev.off()
#system("open fig.png")
```

# What about "married"?

## For different professions

```{r, echo=FALSE}
# PLOTS
par(las=1)
barplot(t(as.matrix(xx[, c("pMm", "pFm")])), #
        beside=TRUE, col=c(colM, colF), names.arg=c("actors", "tennis players", "scientists"), ylim=c(0,1), main="Proportions, by gender, of individuals for whom \n'married' appears as suggestion in a Google search", cex.main=0.85, legend = c("Men", "Women"))
```

### Actors
The difference between male (`r round(xx$nMm[xx$cat=="actors"], 2)`/`r round(xx$nM[xx$cat=="actors"], 2)`) and female (`r round(xx$nFm[xx$cat=="actors"], 2)`/`r round(xx$nF[xx$cat=="actors"], 2)`) actors in terms of the frequency of the "married" suggestion by Google is not significant. 
```{r}
summary(glm(formula = married ~ sex, data=t.actors, family = binomial))
```

### Tennis players
The word "married" occured for `r round(xx$nMm[xx$cat=="tennis"], 2)` male players and `r round(xx$nFm[xx$cat=="tennis"], 2)` female players; the difference is not significant either.
```{r}
summary(glm(formula = married ~ sex, data=t.tennis, family = binomial))
```

### Scientists
For the scientists:
```{r}
summary(glm(formula = married ~ sex, data=t.science, family = binomial))
```
There is also an effect of sex on the occurence of the word "married" in Google suggestions. 
    
## "married" for HHMI + AAAS scientists

```{r}
model.hhmiaaas.m.s <- glm(formula = married ~ sex, data=sciencesubset, family = binomial)
model.hhmiaaas.m.t <- glm(formula = married ~ TED, data=sciencesubset, family = binomial)
model.hhmiaaas.m.st <- glm(formula = married ~ sex + TED, data=sciencesubset, family = binomial)
model.hhmiaaas.m.i <- glm(formula = married ~ sex * TED, data=sciencesubset, family = binomial)

model.sel(model.hhmiaaas.m.s, model.hhmiaaas.m.t, model.hhmiaaas.m.st, model.hhmiaaas.m.i)
```

The models with and without sex*TED interaction are not distinguishable based on AICc, so we will choose the simpler:
```{r}
summary(model.hhmiaaas.m.st)
```

Here are the numbers of AAAS and HHMI scientists for whom "married" appears as suggestion:
```{r echo=FALSE}
dissectscience2 <- tapply(sciencesubset$married, list(TED=sciencesubset$TED, sex=sciencesubset$sex), sum)
dissectscience2
```

In terms of proportions for each sex, this corresponds to
```{r echo=FALSE}
dissectscience2 <- dissectscience2 / matrix(rep(c(sum(sciencesubset$sex==dimnames(dissectscience2)$sex[1]), sum(sciencesubset$sex==dimnames(dissectscience2)$sex[2])), 2), byrow=TRUE, ncol=2)
#dissectscience2 <- dissectscience2[, c("M", "W")] # Invert the two columns
dissectscience2 <- dissectscience2[c("yes", "no"), ] # Invert the two lines

dmod2 <- cbind(c(dissectscience2[,1], 0, 0), c(0, 0, dissectscience2[,2])) # Trick to make R believe that there are four categories, then we can tune the colors as we wish
par(las=1)
barplot(dmod2, beside=FALSE, col=matrix(c(colMT, colM, colFT, colF), ncol=2), border=TRUE, legend=c("Men, TED", "other men", "Women, TED", "Other women"), ylim=c(0, 0.1), names.arg=c("Men", "Women"), main="Proportions, by gender, of AAAS and HHMI scientists for whom \n'married' appears as a suggestion in a Google search", cex.main=0.85)
```
    

# Comments?

If you have comments or suggestions, feel free to [contact me](mailto:florence.debarre@normalesup.org)!
    
    
    
